SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-02-19 paper.tex
lots of changes

@@

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l@{}}
\toprule
CPU Model        & Xeon X5650      & E5-2690 V3     \\
\midrule
Instruction Set  & SSE4.2 	   & AVX2 	    \\
Vector Width     & \SI{128}{\bit}  & \SI{256}{\bit} \\
Cores            & $2 \times 6$    & $2 \times 12$  \\
Identifier       & \sse/ 	   & \avx/  	    \\
OpenCL Version   & \num{1.2}       & \num{1.2}      \\
\bottomrule
\end{tabular}
\caption{The Intel CPU configurations used in this study. The identifier field will be used as a shorthand descriptor in the performance plots to quickly identify the CPU type.}
\label{t:cpus}
\end{table}

\begin{table}
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
NVIDIA Model   & Tesla C2075    & Tesla K40m    \\
\midrule
Driver Version & \num{384.81}   & \num{387.26}  \\
CUDA Cores     & \num{448}      & \num{2880}    \\
Identifier     & \gpuold/ 	& \gpunew/	\\
OpenCL Version & \num{1.1}	& \num{1.2}	\\
\bottomrule
\end{tabular}
\caption{The NVIDIA GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.
Again, the identifier field will be used as a shorthand descriptor during analysis of the performance results.}
\label{t:gpus}
\end{table}

\begin{table}
\centering
\begin{tabular}{@{}l c c c@{}}
\toprule
Platform & Parallel & Shallow Vectorization & Deep Vectorization \\
\midrule
OpenMP & X & -- & -- \\
POCL OpenCL & X & X & X \\
Intel OpenCL & X & X & -- \\
NVIDIA OpenCL & -- & X & -- \\
\bottomrule
\end{tabular}
\caption{The various platforms used in this study and the execution \slash vectorization patterns they are capable of running}
\label{t:platforms}
\end{table}

@@

\begin{table}[htb]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1]@{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & \multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}} & \multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}} & \multicolumn{1}{l}{\ce{iC5H11OH}\cite{Sarathy:2013jr}} \\
\midrule
\multicolumn{1}{c}{$E_{R^{\prime}}$}                    & 1.56e-8 & 2.95e-8 & 9.42e-8 & 4.86e-4 \\
\multicolumn{1}{c}{$E_{R^{\prime\prime}}$}              & 6.92e-8 & 6.53e-8 & 1.20e-7 & 5.07e-4 \\
\multicolumn{1}{c}{$E_{R}$}                             & 1.49e1  & 1.11e0  & 2.80e0  & 4.82e-1 \\
\multicolumn{1}{c}{$E_{\varepsilon}$}                   & 1.48e1  & 1.13e0  & 2.93e0  & 5.03e-1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} n}{\text{d} t}}$} & 2.53e1  & 2.60e0  & 7.62e0  & 1.58e1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} T}{\text{d} t}}$} & 3.94e5  & 3.35e8  & 3.95e6  & 7.11e7 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} S}{\text{d} t}}$} & 3.52e12 & 3.46e12 & 3.44e12 & 3.38e12 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species and temperature rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
The ``S'' in $E_{\frac{\text{d} S}{\text{d} t}}$ refers to the thermodynamic state parameter, either $V$ or $P$ for \conp/ and \conv/ respectively.
}
\label{T:source_error}
\end{table}

@@

\begin{table}[tbp]
\centering
\begin{tabular}{@{}l S S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Sample size}   & \multicolumn{1}{c}{$\bar{\mathcal{T}}$}  & \multicolumn{1}{c}{$E_{10^{20}}$}   & \multicolumn{1}{c}{$E_{10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{900900}                      & \num{6.431e+18}      & \num{8.084e-1}  & \num{1.907e-5} \\
GRI-Mech 3.0          & \num{450900}                      & \num{7.783e+19}      & \num{1.890e-7}  & \num{1.877e-7} \\
USC-Mech II           & \num{91800}                       & \num{2.830e+21}      & \num{1.540e-3}  & \num{1.704e-7} \\
\ce{iC5H11OH}         & \num{450900}                      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all samples.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics..
}
\label{T:error}
\end{table}

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The wrapping code---responsible for initializing\slash transfering memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms, and \texttt{gcc 4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3}'' was used in all cases, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the ``exact'' form of the Jacobian (as opposed the the ``appropriate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless otherwise stated the \conp/ assumption was used (the differences between \conp/ and \conv/ evaluations will be explored) and the runtimes reported are for the number of conditions specified in~\cref{T:error}.

@@

\Cref{F:cpu_source} explores the performance of the source term evaluations generated by \texttt{pyJac} on the CPU test platforms listed in~\cref{t:cpus}.
In~\cref{F:intel_source_noorm}, the mean runtime per initial condition is shown for both the~\avx/ and~\sse/ CPUs, using both Intel OpenCL and OpenMP.
This normalization is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{T:error}).
For both CPUs, the OpenMP implementation is the slowest over all models; interestingly the parallel (non-vectorized) Intel OpenCL code is slightly faster than OpenMP in all cases.\todo{rerun OpenMP w/ mtune=native}
As expected, the \avx/ machine is moderately faster than the \sse/ CPU; additionally, the wide-vectorized OpenCL code is significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

@@

\begin{figure}[htb]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source_nonorm.pdf}
      \caption{The mean runtime per condition for each chemical model using the Intel OpenCL and OpenMP on both CPUs studied.}
      \label{F:intel_source_noorm}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source.pdf}
      \caption{The speedup achieved over the baseline OpenMP parallelization by both the unvectorized and wide-vectorized Intel OpenCL codes.}
      \label{F:intel_source}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{pocl_source_nonorm.pdf}
      \caption{The mean runtime per condition of the Portable OpenCL runtime compared to OpenMP parallelization.}
      \label{F:pocl_source}
  \end{subfigure}
 \caption{Mean runtimes per condition and speedups achieved by the various CPU OpenCL runtimes compared to OpenMP parallelization for each chemical model studied. The names in the legends correspond to the identifiers listed in~\cref{t:cpus}}
 \label{F:cpu_source}
\end{figure}

\Cref{F:gpu_source} shows the performance of the source term evaluation on the GPUs listed in~\cref{t:gpus}.
In~\cref{F:gpu_source_scaling}, the number of initial conditions over which the source terms are evaluated on the \gpunew/ GPU is varied in order to determine the effect on the mean runtime per condition; the mean runtime continues to decrease for all chemical models until \textasciitilde\num{e4} conditions, at which point the GPU becomes saturated and performance levels off.
\Cref{F:gpu_source_speedup} shows the speedup in source term evaluations the \gpunew/ GPU over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling}; the speedup ranges from \textasciitilde\SIrange{1.75}{2.5}{$\times$} depending on the chemical model in question.
Additionally,~\cref{F:gpu_source_speedup} shows that varying the vector width size (a.k.a., the block size in NVIDIA terminology) has minimal effect on performance; it is possible that use of larger vector sizes could have an effect on performance, but this was not considered in this work.

\todo{rerun NVIDIA on srv2 -- see if that high variability in isopentanol persists}
\begin{figure}[htb]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean runtime per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source term evaluation performance on the NVIDIA GPUs}
  \label{F:gpu_source}
\end{figure}

@@

\Cref{F:source_gpu_cvsf} shows the effect of the ``C'' versus ``F''-ordering on source term evaluation performance on both GPUs.
The speedup reported is calculated per-GPU, such that the effect of data ordering can be seen on each individual GPU.
In general, the ``C''-ordered wide-vectorization is \textasciitilde\SI{1.05}{1.2}{$\times$} faster on the \newgpu/ GPU while the ``C'' and ``F''-ordered data are nearly equivalent for the \gpuold/ GPU, except for the isopentanol model where the ``F''-ordered data is \textasciitilde\SI{1.4}{$\times$} faster (though, we note that there was high variablity in this case). \todo{again, pending rerun on \gpuold/}
The ``C''-ordered speedup on the \gpunew/ GPU is counter to what one might expect, as typically speaking coalesced memory access in a wide-vectorization is easier to achieve if ``F''-ordering is used (see~\cref{S:data}).
However, the vectorized-data ordering here ensures that not only are memory accesses aligned to the vector width size (and thus, likely coalesced) but the improved data-locality; coupled with the larger cache sizes of the \gpunew/ GPU result in this modest speedup for ``C''-data ordering.

\begin{figure}[htb]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean runtime per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP for \avx/.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is calculated per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_cvsf.pdf}
      \caption{The affect on performance of ``C'' vs ``F''-ordering for the wide-vectorized NVIDIA OpenCL code on both GPUs with a vector width of 128.  The speedup presented is calculated per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_gpu_cvsf}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, as well as ``C'' and ``F'' data-ordering on source term evaluation performance in \texttt{pyJac}.
	   It is noted the wide-vectorized ``C'' ordered OpenCL case corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}

\end{figure}

>>>>

\DeclareSIUnit\doubles{doubles}

@@

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l@{}}
\toprule
CPU Model        & Xeon X5650      & E5-2690 V3     \\
\midrule
Instruction Set  & SSE4.2 	   & AVX2 	    \\
Vector Width     & \SI{2}{\doubles}&\SI{4}{\doubles}\\
Cores            & $2 \times 6$    & $2 \times 12$  \\
Identifier       & \sse/ 	   & \avx/  	    \\
OpenCL Version   & \num{1.2}       & \num{1.2}      \\
\bottomrule
\end{tabular}
\caption{The Intel CPU configurations used in this study.
	 The vector width reported in this table is reported in (ideal) number of double operations per SIMD instruction, as this will be used in measuring SIMD efficiency; for reference, the vector width in bits of the \sse/ and \avx/ machines are \SI{128}{\bit} and \SI{256}{\bit} respectively.
	 The identifier field will be used as a shorthand descriptor in the performance plots to quickly identify the CPU type.}
\label{t:cpus}
\end{table}

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
NVIDIA Model   & Tesla C2075    & Tesla K40m    \\
\midrule
Driver Version & \num{384.81}   & \num{387.26}  \\
CUDA Cores     & \num{448}      & \num{2880}    \\
Identifier     & \gpuold/ 	& \gpunew/	\\
OpenCL Version & \num{1.1}	& \num{1.2}	\\
\bottomrule
\end{tabular}
\caption{The NVIDIA GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.
Again, the identifier field will be used as a shorthand descriptor during analysis of the performance results.}
\label{t:gpus}
\end{table}

\begin{table}[htb]
\centering
\begin{tabular}{@{}l c c c@{}}
\toprule
Platform & Parallel & Shallow Vectorization & Deep Vectorization \\
\midrule
OpenMP & X & -- & -- \\
POCL OpenCL & X & X & X \\
Intel OpenCL & X & X & -- \\
NVIDIA OpenCL & -- & X & -- \\
\bottomrule
\end{tabular}
\caption{The various platforms used in this study and the execution \slash vectorization patterns they are capable of running}
\label{t:platforms}
\end{table}

@@

\begin{table}[htbp]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}l S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1]@{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & \multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}} & \multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}} & \multicolumn{1}{l}{\ce{iC5H11OH}\cite{Sarathy:2013jr}} \\
\midrule
\multicolumn{1}{c}{$E_{R^{\prime}}$}                    & 1.56e-8 & 2.95e-8 & 9.42e-8 & 4.86e-4 \\
\multicolumn{1}{c}{$E_{R^{\prime\prime}}$}              & 6.92e-8 & 6.53e-8 & 1.20e-7 & 5.07e-4 \\
\multicolumn{1}{c}{$E_{R}$}                             & 1.49e1  & 1.11e0  & 2.80e0  & 4.82e-1 \\
\multicolumn{1}{c}{$E_{\varepsilon}$}                   & 1.48e1  & 1.13e0  & 2.93e0  & 5.03e-1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} n}{\text{d} t}}$} & 2.53e1  & 2.60e0  & 7.62e0  & 1.58e1 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} T}{\text{d} t}}$} & 3.94e5  & 3.35e8  & 3.95e6  & 7.11e7 \\
\multicolumn{1}{c}{$E_{\frac{\text{d} S}{\text{d} t}}$} & 3.52e12 & 3.46e12 & 3.44e12 & 3.38e12 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species and temperature rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
The ``S'' in $E_{\frac{\text{d} S}{\text{d} t}}$ refers to the thermodynamic state parameter, either $V$ or $P$ for \conp/ and \conv/ respectively.
}
\label{T:source_error}
\end{table}

@@

\begin{table}[htbp]
\centering
\begin{tabular}{@{}l S S[table-format=1.3e1] S[table-format=1.3e1] S[table-format=1.3e1]@{}}
\toprule
Model                 & \multicolumn{1}{c}{Sample size}   & \multicolumn{1}{c}{$\bar{\mathcal{T}}$}  & \multicolumn{1}{c}{$E_{10^{20}}$}   & \multicolumn{1}{c}{$E_{10^{15}}$} \\
\midrule
\ce{H2}\slash \ce{CO} & \num{900900}                      & \num{6.431e+18}      & \num{8.084e-1}  & \num{1.907e-5} \\
GRI-Mech 3.0          & \num{450900}                      & \num{7.783e+19}      & \num{1.890e-7}  & \num{1.877e-7} \\
USC-Mech II           & \num{91800}                       & \num{2.830e+21}      & \num{1.540e-3}  & \num{1.704e-7} \\
\ce{iC5H11OH}         & \num{450900}                      & \num{2.733e+26}      & \num{1.363e-3}  & \num{2.764e-5} \\
\bottomrule
\end{tabular}
\caption{Summary of Jacobian matrix validation results.
Error statistics are the largest filtered relative error $E_\mathcal{C}$ for over all samples.
It is noted that the threshold described in~\cref{e:thresh} varies slightly between the \conp/ and \conv/ cases; the reported $\bar{\mathcal{T}}$ is the average of the two, however the appropriate value was used during calculations of the error statistics..
}
\label{T:error}
\end{table}

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The wrapping code---responsible for initializing\slash transfering memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \avx/\slash\gpunew/ platforms, and \texttt{gcc 4.8.5} on the \sse/\slash\gpuold/ machines.
The optimization level ``\texttt{-O3}'' was used in all cases, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the ``exact'' form of the Jacobian (as opposed the the ``appropriate'' form discussed in~\cref{S:sparsity}) was used in all cases.
Finally, unless otherwise stated the \conp/ assumption was used, a vector width of \num{8}\slash\num{128} was used for all CPU\slash GPU cases respectively, ``C''\slash ``F''-ordered data was used on the CPU\slash GPU, and the runtimes reported are for the number of conditions specified in~\cref{T:error} and include data transfer overhead to\slash from internal buffers used in \texttt{pyJac}.
The effects of choice of vector-width on the CPU\slash GPU, as well as the differences between \conp/ and \conv/ evaluations will be explored in~\cref{S:source_results}, and briefly discussed in~\cref{S:jacobian_results}.

@@

\Cref{F:cpu_source} explores the performance of the source term evaluations generated by \texttt{pyJac} on the CPU test platforms listed in~\cref{t:cpus}.
Source term evaluations, critial in their own right for direct numerical simulations~\cite{Spafford:2010aa} among other applications, also give us a convenient platform to detail the effects of various code configuration options before investigating the more involved Jacobian evaluation peformance.

In~\cref{F:intel_source_noorm}, the mean runtime per initial condition is shown for both the~\avx/ and~\sse/ CPUs, using both Intel OpenCL and OpenMP.
This normalization is chosen to account for the varying numbers of conditions in the PaSR databases for each model (\cref{T:error}).
For both CPUs, the OpenMP implementation is the slowest over all models; interestingly the parallel (non-vectorized) Intel OpenCL code is slightly faster than OpenMP in all cases.\todo{rerun OpenMP w/ mtune=native}
As expected, the \avx/ machine is moderately faster than the \sse/ CPU; additionally, the wide-vectorized OpenCL code is significantly faster than either the OpenMP or unvectorized OpenCL codes on both processors.

@@

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source_nonorm.pdf}
      \caption{The mean runtime per condition for each chemical model using the Intel OpenCL and OpenMP on both CPUs studied.}
      \label{F:intel_source_noorm}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{intel_source.pdf}
      \caption{The speedup achieved over the baseline OpenMP parallelization by both the unvectorized and wide-vectorized Intel OpenCL codes.}
      \label{F:intel_source}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{pocl_source_nonorm.pdf}
      \caption{The mean runtime per condition of the Portable OpenCL runtime compared to OpenMP parallelization.}
      \label{F:pocl_source}
  \end{subfigure}
 \caption{Mean runtimes per condition and speedups achieved by the various CPU OpenCL runtimes compared to OpenMP parallelization for each chemical model studied. The names in the legends correspond to the identifiers listed in~\cref{t:cpus}}
 \label{F:cpu_source}
\end{figure}

\Cref{F:gpu_source} shows the performance of the source term evaluation on the GPUs listed in~\cref{t:gpus}.
In~\cref{F:gpu_source_scaling}, the number of initial conditions over which the source terms are evaluated on the \gpunew/ GPU is varied in order to determine the effect on the mean runtime per condition; the mean runtime continues to decrease for all chemical models until \textasciitilde\num{e4} conditions, at which point the GPU becomes saturated and performance levels off.
\Cref{F:gpu_source_speedup} shows the speedup in source term evaluations the \gpunew/ GPU over the \gpuold/ GPU for the maximum number of conditions in~\cref{F:gpu_source_scaling}; the speedup ranges from \textasciitilde\SIrange{1.75}{2.5}{$\times$} depending on the chemical model in question.
Additionally,~\cref{F:gpu_source_speedup} shows that varying the vector width size (a.k.a., the block size in NVIDIA terminology) has minimal effect on performance; it is possible that use of larger vector sizes could have an effect on performance, but this was not considered in this work.\todo{rerun NVIDIA on srv2 -- see if that high variability in isopentanol persists}

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_scaling.pdf}
      \caption{The mean runtime per condition for each chemical model on the \gpunew/ GPU as a function of the number of initial conditions tested.}
      \label{F:gpu_source_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{gpu_source_speedup.pdf}
      \caption{The speedup achieved on the \gpunew/ versus the \gpuold/ GPU; the names correspond to the identifiers listed in~\cref{t:gpus}}
      \label{F:gpu_source_speedup}
  \end{subfigure}
  \caption{\texttt{pyJac} source term evaluation performance on the NVIDIA GPUs}
  \label{F:gpu_source}
\end{figure}

@@

\Cref{F:source_gpu_cvsf} shows the effect of the ``C'' versus ``F''-ordering on source term evaluation performance on both GPUs.
The speedup reported is calculated per-GPU, such that the effect of data ordering can be seen on each individual GPU.
In general, the ``C''-ordered wide-vectorization is \textasciitilde\SIrange{1.05}{1.2}{$\times$} faster on the \gpunew/ GPU while the ``C'' and ``F''-ordered data are nearly equivalent for the \gpuold/ GPU, except for the isopentanol model where the ``F''-ordered data is \textasciitilde\SI{1.4}{$\times$} faster (though, we note that there was high variablity in this case). \todo{again, pending rerun on \gpuold/}
The ``C''-ordered speedup on the \gpunew/ GPU is counter to what one might expect, as typically speaking coalesced memory access in a wide-vectorization is easier to achieve if ``F''-ordering is used (see~\cref{S:data}).
However, the vectorized-data ordering here ensures that not only are memory accesses aligned to the vector width size (and thus, likely coalesced) but the improved data-locality; coupled with the larger cache sizes of the \gpunew/ GPU result in this modest speedup for ``C''-data ordering.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean runtime per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP for \avx/.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is calculated per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_vector_width.pdf}
      \caption{The effect of vector-width on ``C''-ordered wide-vectorized Intel OpenCL source term evaluations in the \avx/ CPU.}
      \label{F:source_vector_width}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_cvsf.pdf}
      \caption{The affect on performance of ``C'' vs ``F''-ordering for the wide-vectorized NVIDIA OpenCL code on both GPUs with a vector width of 128.  The speedup presented is calculated per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_gpu_cvsf}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, as well as ``C'' and ``F'' data-ordering on source term evaluation performance in \texttt{pyJac}.
	   It is noted the wide-vectorized ``C'' ordered OpenCL case corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}
\end{figure}

Finally,~\cref{F:source_scaling} displays the (strong) parallel scaling efficiency and SIMD efficiency for the CPU platforms.
The strong parallel scaling efficiency is defined as:
\begin{equation}
 \label{e:strong_scaling}
 \varepsilon = \frac{\bar{t}_{1}}{N \bar{t}_{N}}
\end{equation}
where $\bar{t}_{N}$ is the mean runtime per condition on $N$ CPU cores, and $\bar{t}_{1}$ the same on a single CPU core.
The strong parallel scaling efficiency measures the speedup due to the use of additional CPU cores, as a fraction of linear speedup; strong scaling tends to decrease with the number of processors used due to memory bandwidth limitations and decreasing computation work allocated per CPU core~\cite{strong_scaling}.

\Cref{F:source_parallel_scaling} shows the strong parallel scaling efficiency of source rate evaluation in \texttt{pyJac} on the \avx/ machine for both the wide-vectorized Intel OpenCL and OpenMP codes.
In general, the \ce{H2}\slash\ce{CO} mechanism has the worst scaling efficiency for Intel OpenCL; although the mechanism scalings vary somewhat for OpenMP, there is no clear trend.
It is likely the poor scaling of the \ce{H2}\slash\ce{CO} for Intel OpenCL results from both its relatively small size, and few falloff\slash chemically activated reaction (and in particular, the additional expensive logarithm and exponential evaluations that accompany them); i.e., there is simply not enough work to fully utilize the SIMD intructions.
Additionally, OpenMP tends to have higher scalings then the wide-vectorized OpenCL code, e.g., \textasciitilde\num{0.8} and \textasciitilde\num{0.7} for \num{4} and \num{8} CPU cores respectively, compared to just \textasciitilde\num{0.7} and \textasciitilde\num{0.5} for OpenCL; although not pictured (to keep the figure readable), the unvectorized Intel OpenCL code has only slightly worse scaling than the OpenMP code, hence the SIMD-vectorization itself the cause of this poorer scaling.
This is again likely due to relatively small amounts of work associated with source rate evaluation; the longest runtime (the isopentanol model on a single core) takes only \textasciitilde$\SI{2}{\sec}$ for OpenCL, and \textasciitilde$\SI{9}{\sec}$ for OpenMP.
By the time 4--8 CPU cores are utilized, the runtime of wide-vectorized OpenCL code can be measured in hundreds of milliseconds, and parallel overhead begins to dominate, decreasing efficiency.
In contrast, the OpenMP code requires \textasciitilde\num{4} cores to run as fast as the wide-vectorized OpenCL code in this case, hence parallel overhead concerns become increasingly important for 8--16 cores.
Indeed, we will see that the strong scalings for sparse Jacobian evaluation---the most computationally intensive task in this work for which the efficiency will investigated---are far more similar between Intel OpenCL and OpenMP.

The SIMD efficiency is defined as:
\begin{equation}
 \label{e:simd_efficiency}
 \varepsilon_{\text{SIMD}} = \frac{\bar{t}_{\text{unvec}}}{W \bar{t}_{\text{wide}}}
\end{equation}
where $\bar{t}_{\text{unvec}}$ is the mean runtime per condition of the unvectorized OpenCL code, $\bar{t}_{\text{wide}}$ the same for the wide-vectorized OpenCL code, and $W$, the vector width reported in number of double operations in~\cref{t:cpus}
This measure compares the actual speedup due to wide-vectorization compared to the nominal vector-width of the machine in question.
\Cref{F:source_simd_scaling} shows the SIMD efficiency of source term evaluation in \texttt{pyJac} on a single core of the \avx/ machine; the larger models have higher SIMD-efficiency, \textasciitilde\num{0.75}, and the smaller models have lower SIMD-efficiency, \textasciitilde\num{0.625}.
This again demonstrates that the SIMD-vectorization becomes efficient as the amount of work to perform increases (with increasing model size).
Interestingly,~\cref{F:source_sse_simd_scaling} shows the SIMD efficiency on the \sse/ machine as greater than one; this is likely caused by a combination of use of an OpenCL vector-width greater than the native CPU vector-width (i.e., \num{8} v.s., \num{2}), and the improved data-locality for the vectorized-data ordering, as discussed in~\cref{S:data}, and results a modest \textasciitilde\SIrange{10}{25}{\percent} improvement over the nominal vector width.

\begin{figure}[htbp]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_parallel_scaling.pdf}
      \caption{The strong parallel scaling efficiency (as defined in~\cref{e:strong_scaling}) of source term evaluation for Intel OpenCL\slash OpenMP on the \avx/ machine.}
      \label{F:source_parallel_scaling}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \avx/ CPU.}
      \label{F:source_simd_scaling}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_sse_simd_efficiency.pdf}
      \caption{The SIMD efficiency (\cref{e:simd_efficiency}) of source term evaluation for the Intel OpenCL runtime on a single core of the \sse/ CPU.}
      \label{F:source_sse_simd_scaling}
  \end{subfigure}
  \caption{The parallel scaling efficiency and SIMD efficiency of source term evaluation for Intel OpenCL on the \avx/ and \sse/ CPUs.}
  \label{F:source_scaling}
\end{figure}

\subsubsection{Jacobian evaluation}
\label{S:jacobian_results}
